\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-06 main.tex
all the  -> \code{CODE} as per jss  guideline

Currently, available spatio-temporal data structure in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses is based on \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts}, both of which has been replaced by
more recent implementations. \pkg{spatstat} \citep{spatstat} implements
a \texttt{ppp} class for point pattern data; and more recent,
\pkg{stars} \citep{stars} implements a spatio-temporal array with the
dplyr's data cube structure \pkg{cubelyr} \citep{cubelyr} as its
backend. While these implementations either store spatial and temporal
variables all in a single table, hence duplicate the spatial variables
for each temporal unit; or split them into two separate tables that has
the problem of manually joining, mentioned in the previously. This
creates a gap in the software development. The requirement for such a
tool is important given the ubiquity of spatio-temporal vector data in
the wild: the Ireland wind data from \pkg{gstat} is an classic example
data that splits variables into its spatial (\texttt{wind.loc}) and
temporal (\texttt{wind}) dimension; Bureau of Meteorology (BoM) provides
climate observations widely applied in agriculture and ecology study.
\newline

This paper describes the implementation of a new spatio-temporal data
structure: \pkg{cubble}. \pkg{cubble} implements a relational data
structure that uses two forms to manage the switch between spatial and
temporal dimension. With this structure, users can manipulate the
spatial or temporal dimension separately, while leaves the linking of
two dimensions to \pkg{cubble}. The software is available from the
Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}. \newline

@@

Spatio-temporal data usually come in various forms and Figure
\ref{fig:illu-input} shows four examples of this. No matter which form
the data is in, these formats share some common components that
characterise the spatio-temporal data. A spatial identifier (\texttt{id}
in the diagram) is the unique identifier of each site. The temporal
identifier (\texttt{t} in the diagram) prescribes the time stamp each
site is recorded. Coordinates, comprising of latitude and longitude
(\texttt{lon} and \texttt{lat} in the diagram), locates each site on the
map. These identifiers will be the building blocks for the data
structure introduced below. Other variables in the data can be
categorised into two groups: spatial variables that are invariant at
each time stamp for every site, i.e.~the name or code of the weather
station, and temporal variables that varies with time.

@@

In a cubble, there are two forms: nested form and long form, and Figure
\ref{fig:illu-cubble} sketches the two forms along with the associated
attributes. The decision on which form to use is output-oriented,
meaning analysts need to first think about whether the output of a
particular operation is identified only by the spatial identifier, or a
combination of spatial and temporal identifier. The nested cubble is
suitable for working with operations that are only identified by the
site and this type of operation can be a pure manipulation of spatial
variables, or a summary of temporal variables by site (i.e.~the output
of counting the number of raining day is only identified by sites and
hence should be performed with the nested form). Underneath the nested
form, a cubble is built from a row-wise dataframe (\texttt{rowwise\_df})
where each site forms a separate group. This structure simplifies the
calculation that involves temporal variables by avoiding the use of
\texttt{map} syntax when working with list-column.

@@

\texttt{as\_cubble()} creates a \pkg{cubble} by supplying the three key
components: \texttt{key} as the spatial identifier; \texttt{index} as
the temporal identifier; and a vector of \texttt{coords} in the order of
longitude first and then latitude. The naming of \texttt{key} and
\texttt{index} follows the convention in the \pkg{tsibble} package. The
cubble created by default is in the nested form. Below is an example of
creating a cubble: \newline

@@

There are a few information in the \pkg{cubble} header: the name of the
\texttt{key} variable (\texttt{id}) and its number (5), bounding box
(\texttt{{[}115.97,\ -32.94,\ 133.55,\ -12.42{]}}), and also the name of
variable nested in the \texttt{ts} column with its type
(\texttt{date\ {[}date{]},\ prcp\ {[}dbl{]},\ tmax\ {[}dbl{]},\ tmin\ {[}dbl{]}}).

@@

The long cubble is suitable to manipulate the time dimension of the
data. The function \texttt{stretch()} switches the nested cubble into
the long cubble by first extracts all the spatial variables into a
separate tibble and store in the \texttt{spatial} attribute and then
unnests the \texttt{ts} column:

@@

Notice here that the third line in the header now shows the name and
type of spatial variables:
\texttt{lat\ {[}dbl{]},\ long\ {[}dbl{]},\ elev\ {[}dbl{]},\ name\ {[}chr{]},\ wmo\_id\ {[}dbl{]}}.

@@

Manipulation on the spatial and temporal dimension can be an iterative
process. Many times, analysts will need to go back and forth between the
nested and long cubble. \texttt{tamp()} inverses the \texttt{stretch()}
that introduced in the previous section, to switch from a long cubble to
the nested cubble:

@@

Some functions require time invariant variables to be in the same table
as the time varying ones. In cubble, this can be done through
\texttt{migrate()}, which moves the time invariant variables into the
long cubble:

@@

Building from an \texttt{tbl\_df} class, \texttt{cubble} has implemented
methods for \texttt{dplyr} generics, which includes:

@@

\pkg{cubble} is also compatible with \pkg{tsibble}. When creating a
cubble from a tsibble, Only the \texttt{coords} argument need to be
specified. In the created nested form, each element of the list-column
\texttt{ts} is of \texttt{tbl\_ts} class so that existing operations on
the tsibble class can still be applied:

\begin{CodeChunk}
\begin{CodeInput}
R> # casting a tsibble into cubble
R> dt <- climate_flat %>% 
+   tsibble::as_tsibble(key = id, index = date) %>% 
+   cubble::as_cubble(coords = c(long, lat))
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts [366 x 4]>
\end{CodeOutput}
\begin{CodeInput}
R> # add station-based features in the nested form.
R> dt %>% 
+   mutate(fabletools::features(ts, tmax, list(tmax_mean = mean, tmax_var = var)))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      tmax_mean tmax_var
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>      <dbl>    <dbl>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_t~      25.7    38.6 
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_t~      26.2    51.1 
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_t~      23.7    45.4 
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_t~      33.1     3.02
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_t~      34.6    24.7 
\end{CodeOutput}
\end{CodeChunk}

@@

Spatial locations can have grouping structure that is either inherent to
the data (i.e.~countries are nested in continents), or formed by
clustering. Rather than analysing variables in the site level,
summarised variables in the cluster level can give a crisper picture of
local areas. In cubble, \texttt{switch\_key()} can be used to create a
new level of grouping of spatial location by specifying a clustering
variable. Figure \ref{fig:illu-hier} illustrates the relationship of
cubbles at station and cluster level, in both the long and nested form.
By specifying
\texttt{cluster\_nested\ \textless{}-\ station\_nested\ \%\textgreater{}\%\ switch\_key(key\ =\ cluster)},
the cubble re-defines the cubble key from \texttt{id} in
\texttt{station\_nested} to \texttt{cluster} in
\texttt{cluster\_nested}. All the spatial variables variant to
\texttt{cluster} are now nested into a \texttt{.val} column and cluster
level variables can be computed in the same fashion as station level
variables in \texttt{station\_nested}. Following the same principal of
\texttt{stretch}, the long form cluster level cubble expands the time
series variables with both indices: \texttt{cluster} and \texttt{id}.
All the cluster level variables are stored separately for recovering the
nested from from the long form.

@@

\texttt{match\_sites()} from \texttt{cubble} matches data based on these
two criteria. It first matches the two data sources spatially through
computing the pairwise distance based on latitude and longitude. Pairs
that pass the spatial matching are then matched temporally through
computing the number of matched peaks within a fixed length window.
Figure \ref{fig:illu-matching} illustrate this temporal matching in more
details. Given two series \texttt{A} and \texttt{a}, 3 peaks have been
picked in each series. An interval, with default length of 5, is
constructed for each peak in series \texttt{A} and the peaks in series
\texttt{a} are tested against whether they fall into the any of the
intervals. In this illustration, there are 2 matches for these two
series. Several arguments are available in \texttt{match\_sites()} to
fine-tune the matching:

@@

\begin{itemize}
\tightlist
\item
  \texttt{spatial\_n\_keep}: the number of spatial match for each site
  to keep
\item
  \texttt{spatial\_dist\_max}: the maximum distance allowed for a
  matched pair
\item
  \texttt{temporal\_n\_highest}: the number of peak used - 3 in the
  example above
\item
  \texttt{temporal\_window}: the length of the interval - 5 in the
  example above
\item
  \texttt{temporal\_min\_match}: the minimum number of matched peak for
  a valid matched pair
\end{itemize}

@@

Cubble fits in naturally with the interactive graphic pipeline discussed
in the literature
\citep{buja1988elements, buja1996interactive, sutherland2000orca, xie2014reactive, cheng2016enabling}.
Diagram \ref{fig:illu-interactive} illustrates how linking works from
the map to the time series with the two forms of a cubble. The map and
time series plot is associated with the nested or long cubble
respectively and when a user action is captured on the map, the site
will be activated in the nested cubble (left). The nested cubble will
communicate to the long cubble to activate all the observations with the
same \texttt{id} (middle). The long cubble will then highlight the
activated series in the time series plot (right).

The linking is also available from the time series plot to the map. The
selection(s) on the time series is through selecting the point(s) on the
time series and once a point is selected, it will be activated in the
long cubble. All the observations that share the same \texttt{id} are
then activated and this includes other points in the same time series in
the long cubble and the corresponding observation of site in the nested
cubble. These activated observations will then being reflected in the
updated plots and Diagram \ref{fig:illu-interactive-2} in the Appendix
illustrates this process.

@@

Glyph map \citep{Wickham2012-yr} plots the time series as single glyph
on the map and in \proglang{R}, \texttt{GGally} implements the glyph map
through the \texttt{glyphs()} function. The function construct a data
frame with calculated position (\texttt{gx}, \texttt{gy}, \texttt{gid})
of each point on the time series using linear algebra (Equation 1 and 2
in \citet{Wickham2012-yr}). The data can then be piped into
\texttt{ggplot} to create the glyph map as:

@@

A re-implementation of the glyph map as a ggproto \texttt{GeomGlyph} has
been made in the \texttt{cubble} package so that the glyph map can be
created with \texttt{geom\_glyph()}:

@@

Polar glyph map can be specify as a parameter, \texttt{polar\ =\ TRUE},
in the \texttt{geom\_glyph()}, along with \texttt{width} and
\texttt{height} in either absolute or relative value. Global and local
scale can be controlled by the parameter \texttt{global\_rescale}, which
default to \texttt{TRUE} for global scaling. Reference box and line can
be added with separate \texttt{geom\_glyph\_box()} and
\texttt{geom\_glyph\_line()}.

@@

Global Historical Climatology Network (GHCN) provides daily climate
measures from stations across the world and the dataset
\texttt{weatherdata::historical\_tmax} extracts the maximum temperature
for 236 Australian stations. All the stations start recording from year
1969 with the longest back to year 1859. The data
\texttt{historical\_tmax} is already presented as a cubble, with
\texttt{id} as the key, \texttt{date} as the index, and
\texttt{c(longitude,\ latitude)} as the coordinates. This example
compares the maximum temperature in two periods: 1971 - 1975 and 2016 -
2020 for stations in Victoria and New South Wales.

@@

Filtering for the period 1971 \textasciitilde{} 1975 and 2016
\textasciitilde{} 2020 is an operation on the time dimension and the
nested cubble needs to be switched to the long cubble by
\texttt{stretch()}:

@@

There are a few stations which don't have records during 1971 - 1975 and
further investigation shows that while the first and last year of each
series is recorded, the missing years in this period is not reported.
These stations are filtered out by examining whether the summarised time
series has a total of 24 months. The long cubble needs to be switched to
the nested form for this spatial operation using \texttt{tamp()}:

@@

Lastly, to create a glyph map, both the major (\texttt{longitude},
\texttt{latitude}) and minor (\texttt{month}, \texttt{tmax}) coordinates
need to be in the same table. Spatial variables can be moved to the long
form with \texttt{migrate()}:

@@

\texttt{tmax} can then be supplied to \texttt{geom\_glyph()} for the
glyph map in Figure \ref{fig:basic-manip} with a station inset on the
top left corner:

@@

This example shows how to organise data at both level with
\texttt{switch\_key()}. The dataset used is
\texttt{weatherdata::climate\_full}, which records daily precipitation
and maximum/ minimum temperature for 639 stations in Australia from 2016
to 2020. A simple kmean algorithm based on the distance matrix is used
here to create 20 clusters. This creates \texttt{station\_nested} as a
station level nested cubble with a cluster column indicating the group
each station belongs to. More complex algorithms can be used for other
problem, as long as there is a mapping from each station to a cluster.

@@

To create a group level cubble, use \texttt{switch\_key()} with the new
key variable, \texttt{cluster}:

@@

With the group level cubble, \texttt{get\_centroid()} is useful to
compute the centroid of each cluster, which will be used as the major
axis for the glyph map later:

@@

From the map, a few water gauges and weather stations are close to each
other and there is the potential that fluctuation of the water level can
reflect the precipitation measured by the climate station. As introduced
in Section 3.2, \texttt{match\_sites()} can be used to match one source
of data with another source in a cubble. Here \texttt{prcp} in
\texttt{climate} will be matched to \texttt{Water\_course\_level} in
\texttt{river}, which is specified in the argument
\texttt{temporal\_var\_to\_match}. \texttt{temporal\_independent}
controls the dataset that will be used to construct the interval (see
the illustration in section 3.2 for details on this). Here the goal is
to see if precipitation will be reflected by the water level in the
river and hence a lagged effect of water level on precipitation. This
will put precipitation data, \texttt{climate}, as the independent. Given
there is one year worth of data, the number of peak
(\texttt{temporal\_n\_highest}) to consider is slightly raised from a
default 20 to 30. \texttt{temporal\_min\_match} filters out pairs don't
have enough match and to return all the pairs, set
\texttt{temporal\_min\_match} to \texttt{0}

@@

The output from matching is also a cubble, with additional column
\texttt{dist} and \texttt{group} produced from spatial matching and
\texttt{n\_match} from the temporal matching.

@@

NetCDF (Network Common Data Form) is a data format commonly used in
climatology community to deliver global mapping of atmosphere, ocean,
and land. It has two major components: dimension to define the
spatiotemporal grid (longitude, latitude, and time) and variable which
populates the defined grid. Attributes are usually associated with
dimension and variable in the NetCDF format data and a
\href{http://cfconventions.org/}{metadata convention for climate and
forecast} has been designed to standardise this data format. Current
packages on manipulating NetCDF data in R include a high-level R
interface: \texttt{ncdf4}\citep{ncdf4}, a low-level interface that calls
C-interface: \texttt{RNetCDF}\citep{rnetcdf, michna2013rnetcdf}, and a
tidyverse implementation: \texttt{tidync}\citep{tidync}.

Cubble provides an \texttt{as\_cubble()} method to coerce the
\texttt{ncdf4} class from the \texttt{ncdf4} package into a
\texttt{cubble}. It maps each combination of longitude and latitude into
an \texttt{id} (the \texttt{key}) and nests time and variables in the
nested form:

@@

Memory limit with Netcdf data in cubble depends on longitude grid point
x latitude grid point x time grid point x number of variable. Cubble can
handle slightly more than 300 x 300 (longitude x longitude) grid points
for 3 variables in one year and spatial grid can be reduced to trade for
longer time period and more variables. A 300 by 300 spatial grid can be
a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or
global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
Subsetting longitude and latitude grid is available through
\texttt{long\_range} and \texttt{lat\_range} if the Netcdf file has
finer resolution than needed.

\begin{CodeChunk}
\begin{CodeInput}
R> # Assume `my_ncdf` has bounding box of [-180, -90, 180, -90] 
R> # at 0.25 degree resolution and subset it to have 
R> # 1 degree resolution:
R> dt <- as_cubble(my_ncdf, vars = c("q", "z"), 
+                 long_range = seq(-180, 180, 1), 
+                 lat_range = seq(-90, 90, 1))
\end{CodeInput}
\end{CodeChunk}

>>>>

Currently, available spatio-temporal data structure in R includes:
\pkg{spacetime} \citep{spacetime}, which proposes four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses is based on \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts}, both of which has been replaced by
more recent implementations. \pkg{spatstat} \citep{spatstat} implements
a \texttt{ppp} class for point pattern data; and more recent,
\pkg{stars} \citep{stars} implements a spatio-temporal array with the
dplyr's data cube structure \pkg{cubelyr} \citep{cubelyr} as its
backend. While these implementations either store spatial and temporal
variables all in a single table, hence duplicate the spatial variables
for each temporal unit; or split them into two separate tables that has
the problem of manually joining, mentioned in the previously. This
creates a gap in the software development. The requirement for such a
tool is important given the ubiquity of spatio-temporal vector data in
the wild: the Ireland wind data from \pkg{gstat} is an classic example
data that splits variables into its spatial (\texttt{wind.loc}) and
temporal (\texttt{wind}) dimension; Bureau of Meteorology (BoM) provides
climate observations widely applied in agriculture and ecology study.

This paper describes the implementation of a new spatio-temporal data
structure: \pkg{cubble}. \pkg{cubble} implements a relational data
structure that uses two forms to manage the switch between spatial and
temporal dimension. With this structure, users can manipulate the
spatial or temporal dimension separately, while leaves the linking of
two dimensions to \pkg{cubble}. The software is available from the
Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

@@

Spatio-temporal data usually come in various forms and Figure
\ref{fig:illu-input} shows four examples of this. No matter which form
the data is in, these formats share some common components that
characterise the spatio-temporal data. A spatial identifier (\code{id}
in the diagram) is the unique identifier of each site. The temporal
identifier (\code{t} in the diagram) prescribes the time stamp each site
is recorded. Coordinates, comprising of latitude and longitude
(\code{lat} and \code{(long)} in the diagram), locates each site on the
map. These identifiers will be the building blocks for the data
structure introduced below. Other variables in the data can be
categorised into two groups: spatial variables that are invariant at
each time stamp for every site, i.e.~the name or code of the weather
station, and temporal variables that varies with time.

@@

In a cubble, there are two forms: nested form and long form, and Figure
\ref{fig:illu-cubble} sketches the two forms along with the associated
attributes. The decision on which form to use is output-oriented,
meaning analysts need to first think about whether the output of a
particular operation is identified only by the spatial identifier, or a
combination of spatial and temporal identifier. The nested cubble is
suitable for working with operations that are only identified by the
site and this type of operation can be a pure manipulation of spatial
variables, or a summary of temporal variables by site (i.e.~the output
of counting the number of raining day is only identified by sites and
hence should be performed with the nested form). Underneath the nested
form, a cubble is built from a row-wise dataframe (\code{rowwise_df})
where each site forms a separate group. This structure simplifies the
calculation that involves temporal variables by avoiding the use of
\code{map} syntax when working with list-column.

@@

`as\_cubble()\} creates a \pkg{cubble} by supplying the three key
components: \code{key} as the spatial identifier; \code{index} as the
temporal identifier; and a vector of \code{coords} in the order of
longitude first and then latitude. The naming of \code{key} and
\code{index} follows the convention in the \pkg{tsibble} package. The
cubble created by default is in the nested form. Below is an example of
creating a cubble:

@@

There are a few information in the \pkg{cubble} header: the name of the
\code{key} variable (\texttt{id}) and its number (5), bounding box
(\texttt{{[}115.97,\ -32.94,\ 133.55,\ -12.42{]}}), and also the name of
variable nested in the \code{ts} column with its type:
\code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

@@

The long cubble is suitable to manipulate the time dimension of the
data. The function \code{stretch()} switches the nested cubble into the
long cubble by first extracts all the spatial variables into a separate
tibble and store in the \code{spatial} attribute and then unnests the
\code{ts} column:

@@

Notice here that the third line in the header now shows the name and
type of spatial variables:
\code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}.

@@

Manipulation on the spatial and temporal dimension can be an iterative
process. Many times, analysts will need to go back and forth between the
nested and long cubble. \code{tamp()} inverses the \code{stretch()} that
introduced in the previous section, to switch from a long cubble to the
nested cubble:

@@

Some functions require time invariant variables to be in the same table
as the time varying ones. In cubble, this can be done through
\code{migrate()}, which moves the time invariant variables into the long
cubble:

@@

Building from an \code{tbl_df} class, \pkg{cubble} has implemented
methods for \pkg{dplyr} generics, which includes:

@@

\pkg{cubble} is also compatible with \pkg{tsibble}. When creating a
cubble from a tsibble, Only the \code{coords} argument need to be
specified:

\begin{CodeChunk}
\begin{CodeInput}
R> # casting a tsibble into cubble
R> dt <- climate_flat %>% 
+   tsibble::as_tsibble(key = id, index = date) %>% 
+   cubble::as_cubble(coords = c(long, lat))
R> dt
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

In the created nested form, each element of the list-column \code{ts} is
of \code{tbl_ts} class so that existing operations on the tsibble class
can still be applied:

\begin{CodeChunk}
\begin{CodeInput}
R> # add station-based features in the nested form.
R> dt %>% 
+   mutate(fabletools::features(ts, tmax, list(tmax_mean = mean, tmax_var = var)))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      tmax_mean tmax_var
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>      <dbl>    <dbl>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_t~      25.7    38.6 
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_t~      26.2    51.1 
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_t~      23.7    45.4 
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_t~      33.1     3.02
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_t~      34.6    24.7 
\end{CodeOutput}
\end{CodeChunk}

@@

Spatial locations can have grouping structure that is either inherent to
the data (i.e.~countries are nested in continents), or formed by
clustering. Rather than analysing variables in the site level,
summarised variables in the cluster level can give a crisper picture of
local areas. In cubble, \code{switch_key()} can be used to create a new
level of grouping of spatial location by specifying a clustering
variable. Figure \ref{fig:illu-hier} illustrates the relationship of
cubbles at station and cluster level, in both the long and nested form.
By specifying
\code{cluster_nested <- station_nested \%>\% switch_key(key = cluster)},
the cubble re-defines the cubble key from \code{id} in
\code{station_nested} to \code{cluster} in \code{cluster_nested}. All
the spatial variables variant to \code{cluster} are now nested into a
\code{.val} column and cluster level variables can be computed in the
same fashion as station level variables in \code{station_nested}.
Following the same principal of \code{stretch}. the long form cluster
level cubble expands the time series variables with both indices:
\code{cluster} and \code{id}. All the cluster level variables are stored
separately for recovering the nested from from the long form.

@@

`match\_sites()\} from \pkg{cubble} matches data based on these two
criteria. It first matches the two data sources spatially through
computing the pairwise distance based on latitude and longitude. Pairs
that pass the spatial matching are then matched temporally through
computing the number of matched peaks within a fixed length window.
Figure \ref{fig:illu-matching} illustrate this temporal matching in more
details. Given two series \code{A} and \code{a}. 3 peaks have been
picked in each series. An interval, with default length of 5, is
constructed for each peak in series \code{A} and the peaks in series
\code{a} are tested against whether they fall into the any of the
intervals. In this illustration, there are 2 matches for these two
series. Several arguments are available in \code{match_sites()} to
fine-tune the matching:

@@

\begin{itemize}
\tightlist
\item
  \code{spatial_n_keep}. the number of spatial match for each site to
  keep
\item
  \code{spatial_dist_max}. the maximum distance allowed for a matched
  pair
\item
  \code{temporal_n_highest}. the number of peak used - 3 in the example
  above
\item
  \code{temporal_window}. the length of the interval - 5 in the example
  above
\item
  \code{temporal_min_match}. the minimum number of matched peak for a
  valid matched pair
\end{itemize}

@@

Cubble fits in naturally with the interactive graphic pipeline discussed
in the literature
\citep{buja1988elements, buja1996interactive, sutherland2000orca, xie2014reactive, cheng2016enabling}.
Diagram \ref{fig:illu-interactive} illustrates how linking works from
the map to the time series with the two forms of a cubble. The map and
time series plot is associated with the nested or long cubble
respectively and when a user action is captured on the map, the site
will be activated in the nested cubble (left). The nested cubble will
communicate to the long cubble to activate all the observations with the
same \code{id} (middle). The long cubble will then highlight the
activated series in the time series plot (right).

The linking is also available from the time series plot to the map. The
selection(s) on the time series is through selecting the point(s) on the
time series and once a point is selected, it will be activated in the
long cubble. All the observations that share the same \code{id} are then
activated and this includes other points in the same time series in the
long cubble and the corresponding observation of site in the nested
cubble. These activated observations will then being reflected in the
updated plots and Diagram \ref{fig:illu-interactive-2} in the Appendix
illustrates this process.

@@

Glyph map \citep{Wickham2012-yr} plots the time series as single glyph
on the map and in \proglang{R}, \pkg{GGally} implements the glyph map
through the \code{glyphs()} function. The function construct a data
frame with calculated position (`gx\}. \code{gy}. \code{gid}. of each
point on the time series using linear algebra (Equation 1 and 2 in
\citet{Wickham2012-yr}). The data can then be piped into \code{ggplot}
to create the glyph map as:

@@

A re-implementation of the glyph map as a ggproto \code{GeomGlyph} has
been made in the \pkg{cubble} package so that the glyph map can be
created with \code{geom_glyph()}:

@@

Polar glyph map can be specify as a parameter, \code{polar = TRUE}. in
the \code{geom_glyph()}. along with \code{width} and \code{height} in
either absolute or relative value. Global and local scale can be
controlled by the parameter \code{global_rescale}. which default to
\code{TRUE} for global scaling. Reference box and line can be added with
separate \code{geom_glyph_box()} and
\textbackslash code\{geom\_glyph\_line()`.

@@

Global Historical Climatology Network (GHCN) provides daily climate
measures from stations across the world and the dataset
\code{weatherdata::historical_tmax} extracts the maximum temperature for
236 Australian stations. All the stations start recording from year 1969
with the longest back to year 1859. The data \code{historical_tmax} is
already presented as a cubble, with \code{id} as the key, \code{date} as
the index, and \code{c(longitude, latitude)} as the coordinates. This
example compares the maximum temperature in two periods: 1971 - 1975 and
2016 - 2020 for stations in Victoria and New South Wales.

@@

Filtering for the period 1971 \textasciitilde{} 1975 and 2016
\textasciitilde{} 2020 is an operation on the time dimension and the
nested cubble needs to be switched to the long cubble by
\code{stretch()}:

@@

There are a few stations which don't have records during 1971 - 1975 and
further investigation shows that while the first and last year of each
series is recorded, the missing years in this period is not reported.
These stations are filtered out by examining whether the summarised time
series has a total of 24 months. The long cubble needs to be switched to
the nested form for this spatial operation using \code{tamp()}:

@@

Lastly, to create a glyph map, both the major (\code{longitude},
\code{latitude}) and minor (\code{month}, \code{tmax}) coordinates need
to be in the same table. Spatial variables can be moved to the long form
with \code{migrate()}:

@@

\code{tmax} can then be supplied to \code{geom_glyph()} for the glyph
map in Figure \ref{fig:basic-manip} with a station inset on the top left
corner:

@@

This example shows how to organise data at both level with
\code{switch_key()}. The dataset used is
\code{weatherdata::climate_full}, which records daily precipitation and
maximum/ minimum temperature for
\code{r nrow(weatherdata::climate_full)} stations in Australia from 2016
to 2020. A simple kmean algorithm based on the distance matrix is used
here to create 20 clusters. This creates \code{station_nested} as a
station level nested cubble with a cluster column indicating the group
each station belongs to. More complex algorithms can be used for other
problem, as long as there is a mapping from each station to a cluster.

@@

To create a group level cubble, use \code{switch_key()} with the new key
variable, \code{cluster}:

@@

With the group level cubble, \code{get_centroid()} is useful to compute
the centroid of each cluster, which will be used as the major axis for
the glyph map later:

@@

From the map, a few water gauges and weather stations are close to each
other and there is the potential that fluctuation of the water level can
reflect the precipitation measured by the climate station. As introduced
in Section 3.2, \code{match_sites()} can be used to match one source of
data with another source in a cubble. Here \code{prcp} in \code{climate}
will be matched to \code{Water_course_level} in \code{river}, which is
specified in the argument \code{temporal_var_to_match}.
\code{temporal_independent} controls the dataset that will be used to
construct the interval (see the illustration in section 3.2 for details
on this). Here the goal is to see if precipitation will be reflected by
the water level in the river and hence a lagged effect of water level on
precipitation. This will put precipitation data, \code{climate}, as the
independent. Given there is one year worth of data, the number of peak
(\code{temporal_n_highest}) to consider is slightly raised from a
default 20 to 30. \code{temporal_min_match} filters out pairs don't have
enough match and to return all the pairs, set \code{temporal_min_match}
to 0.

@@

The output from matching is also a cubble, with additional column
\code{dist} and \code{group} produced from spatial matching and
\code{n_match} from the temporal matching.

@@

NetCDF (Network Common Data Form) is a data format commonly used in
climatology community to deliver global mapping of atmosphere, ocean,
and land. It has two major components: dimension to define the
spatiotemporal grid (longitude, latitude, and time) and variable which
populates the defined grid. Attributes are usually associated with
dimension and variable in the NetCDF format data and a
\href{http://cfconventions.org/}{metadata convention for climate and
forecast} has been designed to standardise this data format. Current
packages on manipulating NetCDF data in R include a high-level R
interface: \pkg{ncdf4} \citep{ncdf4}, a low-level interface that calls
C-interface: \pkg{RNetCDF} \citep{rnetcdf, michna2013rnetcdf}, and a
tidyverse implementation: \pkg{tidync} \citep{tidync}.

Cubble provides an \code{as_cubble()} method to coerce the \code{ncdf4}
class from the \pkg{ncdf4} package into a \code{cubble}. It maps each
combination of longitude and latitude into an \code{id} (the \code{key})
and nests time and variables in the nested form:

@@

Memory limit with Netcdf data in cubble depends on longitude grid point
x latitude grid point x time grid point x number of variable. Cubble can
handle slightly more than 300 x 300 (longitude x longitude) grid points
for 3 variables in one year and spatial grid can be reduced to trade for
longer time period and more variables. A 300 by 300 spatial grid can be
a bounding box of {[}100, -80, 180, 0{]} at 0.25 degree resolution or
global bounding box {[}-180, -90, 180, -90{]} at 1 degree resolution.
Subsetting longitude and latitude grid is available through
\code{long_range} and \code{lat_range} if the Netcdf file has finer
resolution than needed.

\begin{CodeChunk}
\begin{CodeInput}
R> # Assume \code{my_ncdf} has bounding box of [-180, -90, 180, -90] 
R> # at 0.25 degree resolution and subset it to have 
R> # 1 degree resolution:
R> dt <- as_cubble(my_ncdf, vars = c("q", "z"), 
+                 long_range = seq(-180, 180, 1), 
+                 lat_range = seq(-90, 90, 1))
\end{CodeInput}
\end{CodeChunk}

Figure \ref{fig:netcdf} gives an example of reproducing Figure 19 in
\citet{hersbach2020era5} using ERA5 data. ERA5 data
\citep{hersbach2020era5} is the latest reanalysis of global atmosphere,
land surface, and ocean waves from 1950 onwards and is available in the
NetCDF format. It can be directly downloaded from the
\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels?tab=overview}{Copernicus
Climate Data Store (CDS)} website or programmatically via R package
\pkg{ecmwfr} \citep{ecwmfr}. Variable Specific humidity and geopotential
are queried on the 10 hPa pressure level for four dates: 2002-09-22,
2002-09-26, 2002-09-30, and 2002-10-04 and once downloaded, the data can
be read into a cubble using the code above. Readers who are interested
in the details of can refer to \citet{simmons2020global} and
\citet{simmons2005ecmwf}.
